---
title: "Halloween Candy"
author: "Nicholas Pacia"
format: gfm
---
# Import Candy Data

Read the csv file after downloading it to project directory.
```{r}
candy_file <- "candy-data.csv"

candy <- read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy) #number of candy types
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
#skittles wildberry are my favorite candy
candy["Skittles wildberry", ]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```


>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Using skim function
Install package.
```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent appears to be on a different scale from the majority of other variables in the dataset.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

0 represents a no or not chocolate, and 1 represents yes for it is chocolate.

## Histogram Plot

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy)+
  aes(winpercent)+
  geom_histogram(bins = 10, fill = "blue")
#hist(candy$winpercent)
```

>Q9. Is the distribution of winpercent values symmetrical?

The distribution is not symmetrical.

>Q10. Is the center of the distribution above or below 50%?

```{r}
t.test(wp <- candy$winpercent)
median(wp) 
```
The center of distribution, the mean, is above 50.

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(wp[as.logical(candy$chocolate)]) > mean(wp[as.logical(candy$fruity)])

```
Chocolate candy is higher ranked than fruity.

>Q12. Is this difference statistically significant?

```{r}
t.test(wp[as.logical(candy$chocolate)], wp[as.logical(candy$fruity)])
```
p-value is 2.871e-08 which is < 0.05 so there is a statistical significance between chocolate and fruity ratings.

# Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
#install.packages("dplyr")
library(dplyr)
candy_ascending <- candy %>%
  arrange(winpercent)
head(candy_ascending)
```
The five least like candies are the first 5 rows of candy_ascending.

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy_descending <- candy %>%
  arrange(desc(winpercent))
head(candy_descending)
```
The top 5 all time favorites are the first 5 rows in candy_descending.

>Q15. Make a first barplot of candy ranking based on winpercent values

```{r}
ggplot(candy) + 
  aes(x = winpercent, y = rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
barplot <- ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent))+
  geom_col()
```
Now we shall add colours.
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

barplot <- barplot + geom_col(fill = my_cols)
barplot
```

>Q17. What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocolate candy.


>Q18. What is the best ranked fruity candy?

Starbursts is the best ranked fruity candy.

# Looking at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures are the highest winpercent with the lowest pricepercent

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Nik L Nip is the most expensive candy with the least popularity.

# Correlation Structure

```{r}
#install.packages("corrplot")
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```


>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruity and chocolate, pluribus and bar, and all the rest that have red colored dots.

>Q23. Similarly, what two variables are most positively correlated?

winpercent and chocolate.

# PCA

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```
```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16) #adding colors
```
Improve the plot. 
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
We can use ggrepel to label the ggplot.
```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")

```
How about we make an interactive plot.
```{r}
#install.packages("plotly")
library(plotly)

#ggplotly(p) #interactive plot cant be rendered as pdf
```

PC1 contributions plot.
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are picked up strongly. These make sense because these categories are the most distinct so they are the best attributes to cluster different candies because candies within these categories have the least overlap into other candy types.

