---
title: "Class 05: Data Visualization with GGPLOT"
author: "Nicholas Pacia"
format: pdf
---

#Our first ggplot ggplot2 package must be downloaded to computer first

To install any package we use the `install.packages()` command.

Now can I use it? No, we must first call `library(ggplot2)`

```{r}
library(ggplot2)
```

```{r}
head(mpg)
```

First plot of displ vs hwy. All ggplot graphs are made in the same way (data+aes+geom)

```{r}
ggplot(mpg) + aes(displ,hwy) + geom_point()
```

Adding more layers

```{r}
ggplot(mpg) + aes(displ,hwy) + geom_point() +
  geom_smooth(method="lm",se=F, color="green")
```

#Plot of gene expression data First read the data from online

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q. How many genes are in this dataset?

```{r}
nrow(genes)
```
What are column names?
```{r}
colnames(genes)
```
```{r}
ggplot(genes, aes(Condition1,Condition2)) +geom_point()
```
Adding colors: add another aes() mapping of color to the `State` column.
```{r}
ggplot(genes, aes(Condition1,Condition2, color=State)) +geom_point()
```
>Q. How many genes are up regulaed and down regulated?

```{r}
table(genes$State)
```

More customization of plot. Plot saved to object `n`. Use `scale_color_manual` to manually set color values. We can then update our object with the new changes so that they are saved for future plots.
```{r}
n <- ggplot(genes, aes(Condition1,Condition2, color=State)) +geom_point()
n <- n + scale_color_manual(values = c("red", "grey", "blue"))
n
```
Now we need to add titles and descriptors with `labs()`. 
```{r}
n <- n + labs(title = "Gene Expression Changes upon Drug Treatment",
           x="Control (no drug)", y="Drug Treatment")
n
```
